---
title: "Data Quality Problems (and Solutions)"
author: "Andrew Martin"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE, message = FALSE, warning = FALSE, error = FALSE}
knitr::opts_chunk$set(collapse = T)
library(nyschooldata)
```

Overall, [data.nysed.gov](http://data.nysed.gov/) does a lot of things right, and I very much appreciate the general spirit of data accessibility that is happening there.

There are some issues that come up when using the released assessment tables.  This vignette documents what I've bumped into, how I've problem solved, and what might be done.

## Assessment 

Here's the top of the Assessment database:

```{r}
ex <- get_raw_assess_db(2016)
```

```{r}

head(ex) %>% print.AsIs()

```

And here are ten random rows:
```{r}

rand_slices <- runif(10, min = 1, max = ex %>% nrow()) %>% round(0)

dirty_random <- ex %>% 
  dplyr::slice(rand_slices) 

dirty_random %>% print.AsIs()

summary(dirty_random)
```

A couple of things I notice:

- districts are mixed with schools, with no apparent field that tags the level (School, District, County or Statewide Aggregation)

- `L1.PCT`, `L2.PCT` etc all have percent symbols, which means that they won't read cleanly into R.

- `TOTAL.TESTED`, `L1.COUNT`, `L2.COUNT`, are character, not numeric (because of supression codes.)

- test grade level isn't a field unto itself - it needs to be text-processed out of the `ITEM.DESC` field.

- there isn't a unique key/identifier per school / subject / subgroup 

- there are all kinds of extra classes thrown onto this data.frame - an unfortunate side effect of `Hmiscs` `mdb.get` function, but more fundamentally a reflection of the fact that Access databases simply aren't a great way to distribute data.

- for the county and NRC aggregates (eg bedscode == `130000000000`), `total_tested` is reported, but `mean_scale_score` is not.

Odds and Ends:

- the 2016 Assessment database has inconsistent headers.  Instead of using `school_year`, it uses `sy_end_date`.

- the 2015 Assessment database has inconsistent variables.  It includes a `SUM_OF_SCALE_SCORE` field.

`clean_assess_db` solves these problems.  Here's the output of that function:
 (`fetch_assess_db` is a wrapper around that function):
 
```{r}

ex2 <- fetch_assess_db(2016)
  
```

And here are ten random rows:
```{r}

clean_random <- ex2 %>% 
  dplyr::slice(rand_slices) 

clean_random %>% print.AsIs()

```
